%!TEX root = ../thesis.tex

\chapter{Publication Overview}

This thesis is based on five published papers. Minor changes have been done to the papers; layout adjustments to make them fit the format of this thesis, proofs has been moved to the main text from appendix or technical report and typos have been fixed. Each paper will now be summarised by its abstract, contribution and details about publication.

In addition to these five papers, I have published the following papers during my Ph.D. study.\\

\noindent{\usekomafont{disposition}Verification of Timed-Arc Petri Nets}\\
Lasse Jacobsen, Morten Jacobsen, \mhms, and \jiris\\
Published in proceedings of the 37th international conference on Current Trends in Theory and Practice of Computer Science (SOFSEM'11) volume 6543 of LNCS, pages 46--72. Springer-Verlag, 2011\\

\noindent{\usekomafont{disposition}Compositional Verification of Real-time Systems Using Ecdar.}\\
Alexandre David, \kims, Axel Legay, \mhms, Ulrik Nyman, Anders P. Ravn, Arne Skou and Andrzej W\c{a}sowski\\
Published in international journal on Software Tools for Technology Transfer (STTT), vol.14, no. 6, p. 703--720. Springer-Verlag, 2012\\

\noindent{\usekomafont{disposition}TAPAAL 2.0: Integrated Development  Environment for Timed-Arc Petri Nets.}\\ Alexandre David, Lasse Jacobsen, Morten Jacobsen, Kenneth Y. J\o rgensen \mhms, and \jiris\\
\nikolas, \jans, \kims, \mhms, \jiris\\
Published in proceedings of the 18th international Conference on Tools and Algorithms for the Construction and Analysis of Systems (TACAS'12) volume 7214 of LNCS, pages 492--497. Springer-Verlag, 2012\\

\noindent{\usekomafont{disposition}TCTL-preserving translations from timed-arc Petri nets to networks of timed automata.}\\
Joakim Byg, Lasse Jacobsen, Morten Jacobsen, Kenneth Y. J\o rgensen \mhms, and \jiris\\
Published in journal Theoretical Computer Science. Currently, only available on-line, printed version in press. doi: 10.1016/j.tcs.2013.07.011 \\

\newpage

\section*{Paper~\ref{pap:atva}: \atva}
\nikola \jan \kim \\ \mhm \jiri
\vspace{3mm}

\noindent\input{papers/atva/abstract}

\subsubsection*{Contributions}

\begin{itemize}
\item Definition of parametric modal transition systems, an extension of MTS that is general enough to express several features missing in previous extensions of MTS.
\item A comprehensive overview of complexity results for modal refinement checking of PMTSs and its subclasses.
\end{itemize}

\subsubsection*{Publication History}

The paper was accepted and presented at the 9th International Symposium on Automated Technology for Verification and Analysis (ATVA'11) and published in the proceedings of ATVA'11, volume 6996 of LNCS, pages 275--289, Springer-Verlag 2011.


\newpage

\section*{Paper~\ref{pap:memics}: \memics}
\mhm \kim \liu
\vspace{3mm}

\noindent\input{papers/memics/abstract}

\subsubsection*{Contributions}

\begin{itemize}
\item Defined a logic for Boolean modal transition systems and proved that it characterises modal refinement on BMTSs.
\item Presented a way to apply the logic on parametric modal transition systems in order to determine which parameter configurations will result in a BMTS that satisfy the requirement.
\end{itemize}

\subsubsection*{Publication History}

The paper was accepted and presented at the 8th Annual Doctoral Workshop on Mathematical and Engineering Methods in Computer Science 2012 (MEMICS'12) and published in the local proceedings of MEMICS'12.


\newpage

\section*{Paper~\ref{pap:lpar}: \lpar}
\nikola \jan \kim \\ \mhm \jiri
\vspace{3mm}

\noindent\input{papers/lpar/abstract}

\subsubsection*{Contributions}

\begin{itemize}
\item Defined an extension of modal transition systems that include variable time durations of actions and a price-scheme with a one-shot investment price for the hardware and a running cost per time unit for each action.
\item Presented an algorithm to determine the cheapest implementation w.r.t. running expenses, on a fixed budget for hardware investment.  
\end{itemize}

\subsubsection*{Publication History}

The paper was accepted and presented at the 18th International Conference on Logic for Programming, Artificial Intelligence and Reasoning (LPAR'12) and published in the proceedings of LPAR'12, volume 7180 of LNCS, pages 122-137, Springer-Verlag, 2012.




\newpage

\section*{Paper~\ref{pap:petrinets}: \petrinets}
\serge \rolf \mhm
\vspace{3mm}

\noindent\input{papers/petrinets/abstract}

\subsubsection*{Contributions}

\begin{itemize}
\item Defined asynchronously composed I/O-Petri nets and studied a series of 14 communication properties of such systems models.
\item Proven that these communication properties are compositional, meaning that properties are preserved by new compositions.
\item Proven that all 14 communication properties are decidable for asynchronously composed I/O-Petri nets.
\end{itemize}

\subsubsection*{Publication History}

The paper was accepted and presented at the 34th International Conference on 
Application and Theory of Petri Nets and Concurrency (Petri Nets'13) and published in the proceedings of Petri Nets'13, volume 7927 of LNCS, pages 369-388, Springer-Verlag, 2013.


\newpage

\section*{Paper~\ref{pap:tgc}: \tgc}
\serge \rolf \mhm
\vspace{3mm}

\noindent\input{papers/tgc/abstract}

\subsubsection*{Contributions}

\begin{itemize}
\item Extended the asynchronously composed I/O-Petri nets with modalities and studied communication properties and refinements of such modal specifications.
\item Developed a complete methodology for designing asynchronously composed systems using our new formalism.
\item Proven that channel properties are preserved by refinement and composition. Furthermore that refinement is preserved by composition and hiding.
\end{itemize}

\subsubsection*{Publication History}

The paper was accepted and presented at the 8th International Symposium on Trustworthy Global Computing (TGC'13) and published in the local pre-proceedings of TGC'13. Will also be published in the post-proceedings of TGC'13, that will appear as a volume in Springer's LNCS series.