%!TEX root = ../../thesis.tex

\begin{figure}[ht]
\centering
	\begin{tikzpicture}[node distance=100pt]
		\node[state,minimum size=20pt,initial] (s0) {$s$};
		\node[state,minimum size=20pt,right=of s0] (s1) {};
		\node[state,minimum size=20pt,right=of s1] (s2) {};
		\node[state,minimum size=20pt,below=of s2] (s3) {};
		\node[state,minimum size=20pt,below=of s1] (s4) {};
		\node[state,minimum size=20pt,below=of s0] (s5) {$t$};
		\node[below=0.4 of s4] {
			$ \Phi(t)= (\BigClean,s)\ \vee (\SkipClean,s)\ \vee (\SmallClean,s)$};
		
		\draw[arc] 
			(s0) edge node[above] {\Wait} node[below] {$\ci{1,5}$} (s1) 
			(s1) edge node[above] {\DriveBus} node[below] {$\ui{6,10}$} (s2) 
			(s2) edge node[sloped,above] {\SmallClean} node[sloped,below] {$\ci{4,6}$} (s3) 
			(s3) edge node[above] {\Wait} node[below] {$\ci{1,5}$} (s4) 
			(s4) edge node[above] {\DriveBus} node[below] {$\ui{6,10}$} (s5) 
			(s5) edge[dashed,bend left=80] node[sloped,above,yshift=-3] {\SmallClean} node[sloped,below,yshift=2] {$\ci{4,6}$} (s0) 
			(s5) edge[dashed,bend right=70] node[sloped,above,yshift=-1] {\BigClean} node[sloped,below,yshift=2] {$\ci{20,30}$} (s0) 
			(s5) edge[dashed] node[sloped,above,yshift=-3] {\SkipClean} node[sloped,below,yshift=2] {0} (s0) 
		{};
	\end{tikzpicture}
	\caption{Example of Dual-Priced Modal Transition Systems with Time Durations, specification $\Spec$.\label{fig:spec1}}
\end{figure}


Consider the specification $\Spec$ in Figure~\ref{fig:spec1} 
describing the work of a shuttle bus driver.
He drives a bus between a hotel and the
airport. First, the driver 
has to \Wait for the passengers at the hotel. This can take one
to five minutes. Since this behaviour is required to be present in all the
implementations of this specification, it is drawn as a solid arrow and 
called a \emph{must} transition. Then the driver has to \DriveBus the bus to
the airport (this takes six to ten minutes)
where he has to do a \SmallClean, then
\Wait before he can \DriveBus the bus back to the hotel. When he returns he can
do either a \SmallClean, \BigClean or \SkipClean of the bus before he
continues. Here we do not require a particular option to be realised in the
implementations, hence we only draw the transitions as dashed arrows. As these
transitions may or may not be present in the implementations, they are called
\emph{may} transitions. However, here the intention is to require at least one
option be realised. Hence, we specify this using a propositional formula $\Phi$
assigned to the state $t$ over its outgoing transitions as described
in~\cite{benes_et_al:OASIcs:2011:3070,BKLMS:ATVA:11}. After performing one of the
actions, the driver starts over again. Note that next time 
the choice in $t$ may differ.

Observe that there are three types of durations on the transitions. First,
there are \emph{controllable} intervals, written in angle brackets. The meaning
of e.g.~$\langle1,5\rangle$ is that in the implementation we can instruct the
driver to wait for a fixed number of minutes in the range. Second, there are
\emph{uncontrollable} intervals, written in square brackets. 
The interval~$[6,10]$ on
the \DriveBus transition means that in the implementation we cannot fix any
particular time and the time can vary, say, depending on the traffic 
and it is chosen
nondeterministically by the environment. Third, the degenerated case of a
single number, e.g.~$0$, denotes that the time taken is always constant and
given by this number. In particular, a~zero duration means that the transition
happens instantaneously.
\begin{figure}[ht]
	\centering
	\subfloat[Specification $\Spec_1$.\label{fig:spec2}]{
	\begin{tikzpicture}[node distance=80pt]
			\node[state,minimum size=20pt,initial] (s0) {$s$};
			\node[state,minimum size=20pt,right=of s0] (s1) {};
			\node[state,minimum size=20pt,below=of s1,xshift=-50pt] (s2) {};
			
			\draw[arc] 
				(s0) edge node[above] {\Wait} node[below] {$\ci{3,5}$} (s1) 
				(s1) edge node[sloped,below] {\DriveBus} node[sloped,above] {$\ui{6,10}$} (s2) 
				(s2) edge node[sloped,below] {\SmallClean} node[sloped,above] {$\ci{4,6}$}(s0)		{};
		\end{tikzpicture}
	}
	\qquad
	\subfloat[Implementation $\Impl_1$.\label{fig:impl1}]{
		\begin{tikzpicture}[node distance=80pt]
			\node[state,minimum size=20pt,initial] (s0) {$s$};
			\node[state,minimum size=20pt,right=of s0] (s1) {};
			\node[state,minimum size=20pt,below=of s1,xshift=-50pt] (s2) {};
			
			\draw[arc] 
				(s0) edge node[above] {\Wait} node[below] {$5$} (s1) 
				(s1) edge node[sloped,below] {\DriveBus} node[sloped,above] {$\ui{6,10}$} (s2) 
				(s2) edge node[sloped,below] {\SmallClean} node[sloped,above] {$6$}(s0)		{};
		\end{tikzpicture}
	}
	\caption{Examples of Dual-Priced Modal Transition Systems with Time Durations.}
\end{figure}

The system $\Spec_1$ is another specification, 
a \emph{refinement} of $\Spec$, where 
we additionally specify that the driver 
must do a \SmallClean after each \DriveBus.
Note that the \Wait interval has been narrowed. 
The system $\Impl_1$ is an implementation of
$\Spec_1$ (and actually also of $\Spec$) 
where all controllable time intervals have
already been fully resolved to their final single values: 
the driver must \Wait for 5
minutes and do the \SmallClean for 6 minutes. Note that uncontrollable
intervals remain unresolved in the implementations and the time is
chosen by the environment each time the action is
performed. This reflects  the inherent uncontrollable uncertainty of the
timing, e.g.~of a traffic.
\begin{figure}[ht]
\centering
	\begin{tikzpicture}[node distance=100pt]
		\node[state,minimum size=20pt,initial] (s0) {$s$};
		\node[state,minimum size=20pt,right=of s0] (s1) {};
		\node[state,minimum size=20pt,right=of s1] (s2) {};
		\node[state,minimum size=20pt,below=of s2] (s3) {};
		\node[state,minimum size=20pt,below=of s1] (s4) {};
		\node[state,minimum size=20pt,below=of s0] (s5) {$t$};
		% \node[below=0.4 of s4] {$ \Phi(t)= (\BigClean,s)\ \vee (\SkipClean,s)\ \vee (\SmallClean,s)$};
		
		\draw[arc] 
			(s0) edge node[above] {\Wait} node[below] {$\ci{1,5}$} (s1) 
			(s1) edge node[above] {\DriveBus} node[below] {$\ui{6,10}$} (s2) 
			(s2) edge node[sloped,above] {\SmallClean} node[sloped,below] {$6$} (s3) 
			(s3) edge node[above] {\Wait} node[below] {$\ci{1,5}$} (s4) 
			(s4) edge node[above] {\DriveBus} node[below] {$\ui{6,10}$} (s5) 
			(s5) edge[dashed,bend left=80] node[sloped,above,yshift=-3] {\SmallClean} node[sloped,below,yshift=2] {$\ui{4,5}$} (s0) 
			(s5) edge[bend right=70] node[sloped,above,yshift=-1] {\BigClean} node[sloped,below,yshift=2] {$\ci{20,30}$} (s0) 
			% (s5) edge[dashed] node[sloped,above,yshift=-3] {\SkipClean} node[sloped,below,yshift=2] {0} (s0) 
			{};
	\end{tikzpicture}
	\caption{Specification $\Spec_2$.\label{fig:spec3}}
\end{figure}
\begin{figure}[ht]
\centering
	\begin{tikzpicture}[node distance=100pt]
		\node[state,minimum size=20pt,initial] (s0) {$s$};
		\node[state,minimum size=20pt,right=of s0] (s1) {};
		\node[state,minimum size=20pt,right=of s1] (s2) {};
		\node[state,minimum size=20pt,below=of s2] (s3) {};
		\node[state,minimum size=20pt,below=of s1] (s4) {};
		\node[state,minimum size=20pt,below=of s0] (s5) {$t$};
		% \node[below=0.4 of s4] {$ \Phi(t)= (\BigClean,s)\ \vee (\SkipClean,s)\ \vee (\SmallClean,s)$};
		
		\draw[arc] 
			(s0) edge node[above] {\Wait} node[below] {$1$} (s1) 
			(s1) edge node[above] {\DriveBus} node[below] {$\ui{6,10}$} (s2) 
			(s2) edge node[sloped,above] {\SmallClean} node[sloped,below] {$6$} (s3) 
			(s3) edge node[above] {\Wait} node[below] {$1$} (s4) 
			(s4) edge node[above] {\DriveBus} node[below] {$\ui{6,10}$} (s5) 
			% (s5) edge[dashed,bend left=80] node[sloped,above,yshift=-3] {\SmallClean} node[sloped,below,yshift=2] {$\ui{4,5}$} (s0) 
			(s5) edge node[sloped,above,yshift=-1] {\BigClean} node[sloped,below,yshift=2] {$30$} (s0) 
			% (s5) edge[dashed] node[sloped,above,yshift=-3] {\SkipClean} node[sloped,below,yshift=2] {0} (s0) 
			{};
	\end{tikzpicture}
	\caption{Implementation $\Impl_2$.\label{fig:impl2}}
\end{figure}

The system $\Spec_2$ is yet another specification and again a 
refinement of $\Spec$, where the
driver can always do a \BigClean in $t$ and 
possibly there is also an~alternative
allowed here of a \SmallClean. Notice that both \SmallClean intervals have been
restricted and changed to uncontrollable. This means that we give up the
control over the duration of this action and if this transition is implemented,
its duration will be every time chosen nondeterministically in that range.
Finally, $\Impl_2$ is then an implementation of $\Spec_2$ and $\Spec$.





\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[node distance=.2]
		\node[anchor=north west,fill=lightlightblue] at (0,0) (run-cost){ 
			\begin{tabular}{l|r}
				$a\in\Sigma$  	&\; $r(a)$	\\ \hline
				\Wait 			& 8\\
				\DriveBus 		& 10\\
				\SmallClean 	& 6\\
				\BigClean 		& 7\\
				\SkipClean 		& 0
			\end{tabular} 
		};
		\node[right=of run-cost.north east,anchor=north west, fill=lightlightblue] (invest-cost){
			\begin{tabular}{l|r}
				$h\in H$  		&\; $\hc(h)$\\ \hline
				\VacuumCleaner  & 100\\
				\Sponge 		& 5
			\end{tabular}
		};

		\node[above=of run-cost.north west,anchor=south west, fill=lightlightblue] (H-set) {$H=\{\VacuumCleaner,\Sponge\}$};
		\node[below=of run-cost.south west,anchor=north west, fill=lightlightblue] (H-req) {
				$ 
					\Psi(a) = \begin{cases}
									\VacuumCleaner & \text{if } a = \BigClean \\
									\Sponge \vee \VacuumCleaner & \text{if } a= \SmallClean \\
									true & \text{otherwise}
								\end{cases}
				$
		};
	\end{tikzpicture}
\caption{Price Scheme.\label{fig:cost}}
\end{figure}	

Furthermore, we develop a way to model cost of resources. Each action is
assigned a \emph{running price} it costs per time unit, e.g.~\DriveBus costs 10
each time unit it is being performed as it can be  seen in the left table of
Figure~\ref{fig:cost}. In addition, in order to perform an action, some
hardware may be needed, e.g.~a~\VacuumCleaner for the \BigClean and its price
is 100 as can be seen on the right. This \emph{investment price} is paid once
only.

Let us now consider the problem of finding an optimal implementation, so that we
spend the least possible amount of money (e.g.~the pay to the driver) per time
unit while conforming to the specification $\Spec$. 
We call this problem \emph{the
cheapest implementation problem}.
The optimal implementation is to buy a vacuum cleaner if one can afford an
investment of 100 and do the \BigClean every time as long as possible and \Wait
as shortly as possible. (Note that \BigClean is more costly per time unit than
\SmallClean but lasts longer.) This is precisely implemented in $\Impl_2$ and
the (worst-case) average cost per time unit is $\approx7.97$. 
If one cannot afford the
vacuum cleaner but only a sponge, the optimal worst case long run average is
then a bit worse and is implemented by doing the \SmallClean as long as
possible and \Wait now as \emph{long} as possible. This is depicted in $\Impl_1$
and the respective average cost per time unit is $\approx8.10$. 
%For further details, see Example~\ref{ex:prices}.
